---
title: "weed and sleep"
output:
  html_document:
    df_print: paged
---

# Marijuana Use and Sleep Health in the Multicenter AIDS Cohort Study


```{r}
#load libraries
library(tidyverse)
library(lmtest)
library(tableone)
```

```{r}
#load data
library(readr)
combined_data <- read_csv("combined_data.csv")
head(combined_data)
```

## Data Cleaning

The data for medication use prior to sleep is quite messy, as it relied completely on free-text data. I used the code below to do my best to identify relatively common sleep-aides or sleep-altering medicines. One thing that I didn't do is use any type of string distance measurement to capture typos and match those to categories. That would be a bit more of a lift so I thought what I have below is good enough for now. 
```{r}
#text data for meds is dirty and needs to be cleaned
cols <- c("SLPMED1", "SLPMED2", "SLPMED3", "SLPMED4", "SLPMED5", "SLPMED6", "SLPMED7", "SLPMED8", "SLPMED9", "SLPMED10", "SLPMED11", "SLPMED12", "SLPMED13")

# Convert to lowercase
combined_data[cols] <- lapply(combined_data[cols], str_to_lower)

# Trim whitespace
combined_data[cols] <- lapply(combined_data[cols], function(x) str_trim(x, side = "both"))

combined_data %>%
  filter(str_detect(SLPMED1, " "))
```
```{r}
#identify sleep aid meds
# Define your meds of interest (assumes lowercase already applied)
sleep_meds_otc <- c("melatonin", "ramelteon")

#prescription sleep meds
sleep_meds_prescription <- c("remeron", "mirtazapine", "trazadone", "seroquel", "quetiapine", "desyrel", "ambien", "zolpidem", "zaleplon", "sonata", "lunesta")

#benzo
benzo <- c("temazepam", "alprazolam", "xanax", "restoril", "clonazepam", "klonopin", "lorazepam", "ativan", "flurazepam")

#antihistamines
antihistamine <- c("diphenhydramine", "benadryl", "hydroxyzine", "atarax", "doxylamine", "unisom")

# Create new column
combined_data$sleep_aid <- apply(combined_data[cols], 1, function(row) {
  if (any(grepl(paste(sleep_meds_otc, collapse = "|"), row, ignore.case = TRUE))) {
    return(1)
  } else if (any(grepl(paste(sleep_meds_prescription, collapse = "|"), row))) {
    return(2)
  } else if (any(grepl(paste(antihistamine, collapse = "|"), row))) {
    return(3)
  } else if (any(grepl(paste(benzo, collapse = "|"), row))) {
    return(4)
  } else return(5)
})

# Make it a factor (optional but useful for modeling/plotting)
combined_data$sleep_aid <- factor(combined_data$sleep_aid, levels = c(1:5),
                                   labels = c("OTC Sleep Aid", "Prescription sleep aid", "antihistamine", "benzo", "None"))

combined_data %>%
  mutate(n = 1) %>%
  group_by(sleep_aid) %>%
  summarise(use = sum(n))

```

```{r}
#pull in qualitative data
#import qol data

# Join the datasets by MACSID and VISIT, keeping only matching rows
merged_data <- inner_join(combined_data, x19024_qol, by = c("MACSID", "VISIT"))


head(merged_data$EMTWBR)

```
### create variables about marijuana use

MACS asks about marijuana use with a factor variable about average marijuana use since last visit ranging from not at all to daily. To calculate average use, I assigned a number of days to each of the factor levels and then averaged that over either two or four years to determine the average number of days per month of marijuana use over either two (twopotuse) or four (longpotuse) years
```{r}
#create marijuana variables

#pull in marijuana data
library(readr)
pot_use <- read_csv("~/Downloads/x19024_pot_use(in).csv")
View(pot_use)

#rename twopotuse and longpotuse because I messed this up before

merged_data %>%
  rename(old_twopotuse = twopotuse) %>%
  rename(old_longpotuse = longpotuse)

#create new variable, average potuse

pot_use <- pot_use %>%
  mutate(avgpotuse = case_when(
    HASHF == 0 ~ 0,
    HASHF == 1 ~ 30.5,
    HASHF == 2 ~ 4.36,
    HASHF == 3 ~ 1,
    HASHF == 4 ~ 0.33,
    is.na(HASHF) ~ NA
  ))

#summarize average pot use, should go from 0 to 30.5
pot_use %>%
  summarise(mean = mean(avgpotuse, na.rm = TRUE), max = max(avgpotuse, na.rm = TRUE))

#determine pot use over two years

pot_use_2 <- pot_use %>% 
  filter(VISIT %in% c(680, 690, 700, 710)) %>%
  filter(!is.na(avgpotuse)) %>%
  group_by(MACSID) %>%
  summarise(twopotuse = mean(avgpotuse))

head(pot_use_2)

pot_use_2 %>%
  summarise(mean = mean(twopotuse), max = max(twopotuse))

#determine pot use over 4.5 years (longpotuse)
pot_use_3 <- pot_use %>% 
  filter(VISIT %in% c(630, 640, 650, 660, 670, 680, 690, 700, 710)) %>%
  filter(!is.na(avgpotuse)) %>%
  group_by(MACSID) %>%
  summarise(longpotuse = mean(avgpotuse))

head(pot_use_3)

pot_use_3 %>%
  summarise(mean = mean(longpotuse), max = max(longpotuse))

#join both of these to the data

new_merged_data <- inner_join(merged_data, pot_use_2, by = "MACSID")

newer_merged_data <- inner_join(new_merged_data, pot_use_3, by = "MACSID")
```

```{r}
#data cleaning for other variables

newer_merged_data <- newer_merged_data %>%
  rename(longpotuse = longpotuse.y) %>%
  rename(twopotuse = twopotuse.y)

weed_data <- newer_merged_data %>%
  select(MACSID, RACE, SEROSTAT, EDUCA2, INCOMNEW, SMOKE, longpotuse, twopotuse, slplat, remlat, SLPWEED, remdur, n3dur, n1dur, n2dur, tstmin, ESSSCORE, PSQIRATE, EMTWBR, sleep_aid)

glimpse(weed_data)

#factor variables to recode
cols <- c("RACE", "SEROSTAT", "EDUCA2", "INCOMNEW", "SMOKE", "SLPWEED")

weed_data[cols] <- lapply(weed_data[cols], as.factor)
str(weed_data)
```
### Missing Data

As is my experience with MACS data, there is a fair amount of missing data for some of the qualitative things we are interested in. 22% of respondents did not have data about marijuana use over the past 2 years (twopotuse) and that number increases to 44% over 4 years (longpotuse). Since those are averages and one's marijuana consumption most likely doesn't change that much over time, we could impute those values. 

Missingness for the rest of the variables actually looks pretty good, including for marijuana use prior to sleep (SLPWEED). 



```{r}
#visualize missing data

#examine this dataset for missing data
sapply(weed_data, function(x) sum(is.na(x)))

#bar graph of missing values
freqDf <- data.frame(table(is.na(weed_data)))

# barplot for visualization

barplot(freqDf$Freq , main = "Total Missing values",
xlab = "Missing Data", ylab = "Frequency",

names.arg = c("FALSE","TRUE"),
col = c("#80dfff","lightgreen"))

# legend for barplot

legend("topright",
c("Non-Missing Values","Missing Values"),
fill = c("#80dfff","lightgreen"))

#install visdat
install.packages("visdat")
library(visdat)

vis_miss(weed_data)

#fair amount is missing for twopotuse and longpotuse

weed_data %>%
  select(twopotuse, longpotuse) %>%
  mutate(n = 1) %>%
  summarise(
    two_missing = sum(is.na(twopotuse)),
    long_missing = sum(is.na(longpotuse)), 
    percent_missing_two = two_missing/sum(n),
    percent_missing_long = long_missing/sum(n)
  )

#so missing 22% two-year and 44% longer
```

What we see below is that for when marijuana use was missing over 2 or 4 years compared to not, the medians of things we are interested in were more or less unchanged, so that is encouraging that our data seems to be misssing more or less at random. 

For missing data on use prior to sleep, there was a little bit more there in terms of differences, but those differences don't hold up to a t-test. 

```{r}
#patterns of missingness

weed_data %>%
  select(twopotuse, longpotuse, slplat, remlat, remdur, tstmin, ESSSCORE, PSQIRATE) %>%
  mutate(miss_twopot = is.na(twopotuse)) %>%
  group_by(miss_twopot) %>%
  summarise(across(everything(), median, na.rm = TRUE))

#overall seems pretty similar for twopotuse

weed_data %>%
  select(twopotuse, longpotuse, slplat, remlat, remdur, tstmin, ESSSCORE, PSQIRATE) %>%
  mutate(miss_longpot = is.na(longpotuse)) %>%
  group_by(miss_longpot) %>%
  summarise(across(everything(), median, na.rm = TRUE))

#and for longpotuse

#use prior to sleep
weed_data %>%
  select(SLPWEED, twopotuse, longpotuse, slplat, remlat, remdur, tstmin, ESSSCORE, PSQIRATE) %>%
  mutate(miss_potuse = is.na(SLPWEED)) %>%
  group_by(miss_potuse) %>%
  summarise(across(everything(), median, na.rm = TRUE))

# a little bit more there in terms of difference
weed_missing <- weed_missing %>%
  mutate(miss_potuse = is.na(SLPWEED))

t.test(remlat~miss_potuse, weed_missing)

t.test(tstmin~miss_potuse, weed_missing)

```

```{r}
#looking at the missing values for categorical variables

weed_missing <- weed_data %>%
  mutate(miss_longpot = is.na(longpotuse),
         miss_twopot = is.na(twopotuse))

myVars <- c("RACE", "miss_twopot", "SEROSTAT",  "EDUCA2",  "INCOMNEW", "SMOKE")

catVars <- c("RACE", "miss_twopot", "SEROSTAT",  "EDUCA2",  "INCOMNEW", "SMOKE")

missing_table_1 <- CreateTableOne(vars = myVars, weed_missing, factorVars = catVars, strata = c("miss_twopot"))
missing_table_1

myVars <- c("RACE", "miss_longpot", "SEROSTAT",  "EDUCA2",  "INCOMNEW", "SMOKE")

catVars <- c("RACE", "SEROSTAT",  "EDUCA2",  "INCOMNEW", "SMOKE")

missing_table_2 <- CreateTableOne(vars = myVars, weed_missing, factorVars = catVars, strata = c("miss_longpot"))
missing_table_2

#overall looks good with slight education difference with longpot and more HIV positive people missing

```

Looking at missingness by categorical variables, we don't really see any differences there other than with education. Again, if this were an issue I think it would be reasonable to correct with imputation and we could always add that in for sensitivity analyses. 

### Missing Data for depression/anxiety

Where we have our biggest missing data issue is with our mental health measures. Right now what I have is the 36-item short form health survey. This was recommended to use on the original concept sheet because people thought that things like pain might contribute to poor sleep and we should have a way to capture that. 

That's all well and good, and there is evidence of correlating scores from the SF short form to other measures of depression and anxiety. The issue is that there's quite a bit of missing data for the emotional well-being sub-score that we would need. 

```{r}
#missing data for emotional wellbeing
weed_data %>%
  select(EMTWBR) %>%
  mutate(n = 1) %>%
  summarise(
    missing = sum(is.na(EMTWBR)),
    percent_missing = missing/sum(n)
  )
```
62% of people didn't have full responses here, yikes! Although it looks like there are not a lot of differences based on missing or not other than potentially some differences by race. 

```{r}
#emotional well-being yes vs no and outcomes of interest

weed_data %>%
  select(EMTWBR, slplat, remlat, remdur, tstmin, ESSSCORE, PSQIRATE) %>%
  mutate(missing = is.na(EMTWBR)) %>%
  group_by(missing) %>%
  summarise(across(everything(), median, na.rm = TRUE))
```

```{r}
#missing data for emotional well-being and categorical variables

weed_missing <- weed_data %>%
  mutate(missing = is.na(EMTWBR))

myVars <- c("RACE", "missing", "SEROSTAT",  "EDUCA2",  "INCOMNEW", "SMOKE")

catVars <- c("RACE", "missing", "SEROSTAT",  "EDUCA2",  "INCOMNEW", "SMOKE")

missing_table_3 <- CreateTableOne(vars = myVars, weed_missing, factorVars = catVars, strata = c("missing"))
missing_table_3
```



## Data Visualization

What we see in the table one is a fair proportion of people using marijuana before their sleep study (about 10%), but relatively low use averaged over 2 and 4 years, with patients using marijuana, on average, one half day per month. 

### Table 1

```{r}
#table 1

myVars <- c("race_collapsed", "SEROSTAT",  "schooling",  "AGENOX", "SLPWEED", "twopotuse", "slplat", "remlat", "remdur", "waso", "PSQIRATE", "on_antidepressant")

catVars <- c("race_collapsed", "SEROSTAT",  "schooling", "on_antidepressant")

table_1 <- CreateTableOne(vars = myVars, data = complete_cases, factorVars = catVars)
table_1

```
### Table 1 Stratified by Marijuana use prior to sleep

There are some differences here; in some of the demographic characeteristics as well as with rem latency

```{r}
table_1_weed_strat <- CreateTableOne(vars = myVars, data = weed_data, factorVars = catVars, strata = "SLPWEED")
table_1_weed_strat
```



### Table 1 Stratified by HIV status

When stratified by HIV status, there's a small difference in sleep latency and a slightly higher proportion of current smokers, though no difference in marijuana use prior to sleep.
```{r}
#stratified by HIV status
table_1_strat <- CreateTableOne(vars = myVars, data = weed_data, factorVars = catVars, strata = "SEROSTAT")
table_1_strat
```
### How much marijuana are the users using?
Looking at average marijuana usage, I was thinking that there are a lot of people who most likely didn't use any marijuana over a 2 year period and that is bringing the average down. In fact, 483 (65%) of the cohort didn't use any at all. 

When looking specifically at patients who reported marijuana use, usage was a little bit higher at 9 days per month


```{r}

weed_data %>%
  filter(twopotuse == 0) %>%
  count()


weed_data %>%
  filter(twopotuse > 0) %>%
  summarise(median_twopotuse = median(twopotuse),
            mean_twopotuse = mean(twopotuse, na.rm = TRUE),
            max = max(twopotuse))
```




```{r}
#data visualization by weed use
weed_data %>%
  group_by(SLPWEED) %>%
  summarise(med_lat = median(slplat),
            med_rem_lat = median(remlat, na.rm = TRUE), 
            med_tst = median(tstmin),
            med_remdur = median(remdur),
            med_esss = median(ESSSCORE))

tst_plot <- weed_data %>%
  ggplot(aes(x = SLPWEED, y = tstmin)) +
  geom_boxplot()

remlat_plot <- weed_data %>%
  ggplot(aes(x = SLPWEED, y = remlat)) +
  geom_boxplot()

remdur_plot <- weed_data %>%
  ggplot(aes(x = SLPWEED, y = remdur)) +
  geom_boxplot()



weed_data %>%
  filter(is.na(SLPWEED) == FALSE) %>%
  pivot_longer(cols = c(remdur, n1dur, n2dur, n3dur),
               names_to = "Stage",
               values_to = "Duration") %>%
  group_by(SLPWEED, Stage) %>%
  summarise(
    mean_duration = mean(Duration, na.rm = TRUE),
    se = sd(Duration, na.rm = TRUE) / sqrt(sum(!is.na(Duration))),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Stage, y = mean_duration, fill = SLPWEED)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = mean_duration - se, ymax = mean_duration + se),
                position = position_dodge(width = 0.9),
                width = 0.2) +
  labs(x = "Sleep Stage", y = "Average Duration", fill = "SLPWEED") +
  theme_minimal()



```





```{r}
#visulization, twopotuse

weed_data %>%
  group_by(twopotuse) %>%
  summarise(med_lat = median(slplat),
            med_rem_lat = median(remlat), 
            med_tst = median(tstmin),
            med_remdur = median(remdur),
            med_esss = median(ESSSCORE))

ggplot(weed_data, aes(x = twopotuse, y = slplat)) +
  geom_line()

ggplot(weed_data, aes(x = twopotuse, y = tstmin)) +
  geom_line()
```
## Data Analysis

### Relationship between use before sleep and outcomes
```{r}
#modeling

#create gaussian function
gaussian_model <- function(x) {
  model <- glm(x ~ SLPWEED + RACE + SEROSTAT + SMOKE + EDUCA2 + sleep_aid + EMTWBR, data = weed_data, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))
}

#sleep latency

 sleep_latency <- gaussian_model(weed_data$slplat)

#rem latency
 
 rem_latency <- gaussian_model(weed_data$remlat)

#tst
tst <- gaussian_model(weed_data$tstmin)

# no differences but very wide confidence intervals; will try some recoding to clean up
```

```{r}
#collapsing some categorical variables

library(forcats)

weed_data <- weed_data %>%
  mutate(race_collapsed = fct_collapse(RACE,
                                       White = 1,
                                       Black = 2,
                                       Hispanic = c(2, 4, 8),
                                       Other = c(5, 6, 7)),
         schooling = fct_collapse(EDUCA2,
                                  'Less than HS' = c(1, 2),
                                  'HS or some college' = c(3, 4, 5),
                                  'College or more' = c(6, 7, 8)),
         sleep_aid_collapsed = fct_collapse(sleep_aid,
                                            'Used Sleep Aid' = c("OTC Sleep Aid", "Prescription sleep aid", "antihistamine", "benzo"),
                                            'Did not use' = "None"))

```

### Relationship between Use before sleep and outcomes with collapsed categoricals
```{r}
#try modeling again

gaussian_model <- function(x) {
  model <- glm(x ~ SLPWEED + race_collapsed + SEROSTAT + SMOKE + schooling + sleep_aid_collapsed + EMTWBR, data = weed_data, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))
}
#sleep latency

 sleep_latency <- gaussian_model(weed_data$slplat)

#rem latency
 
 rem_latency <- gaussian_model(weed_data$remlat)

#tst
tst <- gaussian_model(weed_data$tstmin)

psqi <- gaussian_model(weed_data$PSQIRATE)

esss <- gaussian_model(weed_data$ESSSCORE)

#still getting some decently wide confidence intervals
```
### relationship between longitudinal pot use and outcomes
```{r}
#look at longitudinal pot use
gaussian_model_twopot <- function(x) {
  model <- glm(x ~ twopotuse + race_collapsed + SEROSTAT + SMOKE + schooling + sleep_aid_collapsed + EMTWBR, data = weed_data, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))
}
#sleep latency

 sleep_latency_twopot <- gaussian_model_twopot(weed_data$slplat)

#rem latency
 
 rem_latency_twopot <- gaussian_model_twopot(weed_data$remlat)

#tst
tst_twopot <- gaussian_model_twopot(weed_data$tstmin)

#psqi
psqi_twopot <- gaussian_model_twopot(weed_data$PSQIRATE)

```

### interaction between before sleep and long-term use

Looks like there are 215 people who reported use over two years who did not use marijuana before sleep on the night of their sleep study

```{r}
#any interaction between twopot and immediate pot

#cross tabs between chronic pot use and use before sleep

weed_data %>%
  mutate(chronic_pot = ifelse(twopotuse > 0, 1, 2)) %>%
  group_by(chronic_pot) %>%
  count(SLPWEED)

weed_data %>%
  mutate(chronic_pot = ifelse(twopotuse > 0, 1, 2)) %>%
  group_by(chronic_pot) %>%
  summarise(mean_use = mean(twopotuse))

gaussian_model_interaction <- function(x) {
  model <- glm(x ~ SLPWEED + twopotuse + SLPWEED*twopotuse + race_collapsed + SEROSTAT + SMOKE + schooling + sleep_aid_collapsed + EMTWBR, data = weed_data, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))
}
#sleep latency

 sleep_latency_int <- gaussian_model_interaction(weed_data$slplat)

#rem latency
 
 rem_latency_int <- gaussian_model_interaction(weed_data$remlat)

#tst
tst_int <- gaussian_model_interaction(weed_data$tstmin)

#psqi
psqi_int <- gaussian_model_interaction(weed_data$PSQIRATE)
```
```{r}
#what about use and n1 duration (will take out emotional wellbeing for now)

model <- glm(n1dur ~ SLPWEED + race_collapsed + SEROSTAT + SMOKE + schooling + sleep_aid_collapsed + EMTWBR, data = weed_data, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))
```

```{r}

model <- glm(seslponset ~ SLPWEED + race_collapsed + SEROSTAT + SMOKE + schooling + sleep_aid_collapsed + EMTWBR, data = weed_data, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))

```

### use before sleep when limited to chronic users

interesting; if you look only at chronic users, there is a relationship between use before sleep and rem latency (in unadjusted model at least)
```{r}
weed_data_chronic <- weed_data %>%
  mutate(chronic_pot = ifelse(twopotuse > 0, 1, 2)) %>%
  filter(chronic_pot == 1) %>%
  filter(!is.na(SLPWEED))

#can't get this to run and i suspsect it's due to small samples in some cells
chronic_use_model <- gaussian_model(weed_data_chronic$remlat)

#try breaking it down
model <- glm(remlat ~ SLPWEED, data = weed_data_chronic, family = "gaussian")
model_ci <- confint(model)
model_results <- exp(cbind(OR = coef(model), model_ci))

#psqi
model_2 <- glm(PSQIRATE ~ SLPWEED, data = weed_data_chronic, family = "gaussian")
model_ci_2 <- confint(model_2)
model_results_2 <- exp(cbind(OR = coef(model_2), model_ci_2))
```


```{r}
#try to make wake after sleep onset variable
new_weed_data <- newer_merged_data %>%
  select(MACSID, RACE, SEROSTAT, EDUCA2, INCOMNEW, SMOKE, longpotuse, twopotuse, slplat, remlat, SLPWEED, remdur, n3dur, n1dur, n2dur, tstmin, ESSSCORE, PSQIRATE, EMTWBR, sleep_aid, recorddurm, seslponset)

#?sleep efficiency = sleep efficiency in sleep onset to wake
#seslponset if so equals sleep efficiency

#to calculate waso WASO = (1 – Sleep Efficiency)*(Total Recording Time) – Sleep Latency
                 #Or WASO = (1-Sleep Efficiency)/(Sleep Efficiency)* [Total Sleep Time]– Sleep Latency

#1-

#total recording time = recorddurmin
#sleep latency = slplat
#total sleep time = tstmin


```

Brett's interest: 

How do things change among the high users
-histogram of chronic use by use on the sleep study night
-more psqi and esss visualization
-can look more into psqi dimensions, particularly sleep quality
-how does pain affect that

```{r}
#table for Sanjay to figure out sleep variablels
new_weed_data %>%
  select(tstmin, recorddurm, slplat, seslponset) %>%
  head(n = 20)

```
### Identify antidepressant use

```{r}
#antidepressant use
ssri <- c("citalopram", "celexa", "escitalopram", "lexapro", "fluoxetine", "prozac", "paroxetine", "paxil", "sertraline", "zoloft")

snri <- c("desvenlafaxine", "pristiq", "duloxetine", "cymbalta", "venlafaxine", "effexor")

tca <- c("amitriptyline", "elavil", "clomipramine", "anafranil", "desipramine", "norpramine", "doxepin", "sinequan", "imipramine", "tofranil", "nortriptyline", "pamelor")

maoi <- c("isocarboxazid", "marplan", "phenelzine", "nardil", "tranylcypromine", "parnate", "selegiline", "emsam", "rasagiline", "azilect")

#text data for meds is dirty and needs to be cleaned
cols <- c("SLPMED1", "SLPMED2", "SLPMED3", "SLPMED4", "SLPMED5", "SLPMED6", "SLPMED7", "SLPMED8", "SLPMED9", "SLPMED10", "SLPMED11", "SLPMED12", "SLPMED13")

# Convert to lowercase
newer_merged_data[cols] <- lapply(new_merged_data[cols], str_to_lower)

# Trim whitespace
newer_merged_data[cols] <- lapply(newer_merged_data[cols], function(x) str_trim(x, side = "both"))

# Create new column
newer_merged_data$antidepressant_use <- apply(newer_merged_data[cols], 1, function(row) {
  if (any(grepl(paste(ssri, collapse = "|"), row, ignore.case = TRUE))) {
    return(1)
  } else if (any(grepl(paste(snri, collapse = "|"), row))) {
    return(2)
  } else if (any(grepl(paste(tca, collapse = "|"), row))) {
    return(3)
  } else if (any(grepl(paste(maoi, collapse = "|"), row))) {
    return(4)
  } else return(5)
})

# Make it a factor (optional but useful for modeling/plotting)
newer_merged_data$antidepressant_use <- factor(newer_merged_data$antidepressant_use, levels = c(1:5),
                                   labels = c("ssri", "snri", "tca", "maoi", "None"))

newer_merged_data %>%
  mutate(n = 1) %>%
  group_by(antidepressant_use) %>%
  summarise(use = sum(n))
```

That proportion of use seems really low and I'm worried that's because we are only capturing use before sleep that people would have had to manually enter

### merge in final data

will merge in antidepressant use that Deanna identified; is more broad then we need but I think better to be more comprehensive

```{r}
#load required package
install.packages("haven")
library(haven)

antidepressant <- read_sas("poly_data_analysis (1).sas7bdat")

str(antidepressant)

antidepressant %>%
  distinct(VISIT)

# sadly this only goes up to visit 65 which is too early

antidepressant %>%
  filter(VISIT == 650) %>%
  count(AD)

```


### WASO

```{r}
#create waso variable

#calcualte sleep period time

newer_merged_data <- newer_merged_data %>%
  mutate(spt = ((tstmin*100)/seslponset)+0.5) %>%
  mutate(waso = (spt - tstmin))

#visualize waso

newer_merged_data %>%
  ggplot(aes(waso)) +
  geom_histogram()

#stratify by slpweed
newer_merged_data %>%
  filter(!is.na(SLPWEED)) %>%
  group_by(SLPWEED) %>%
  ggplot(aes(waso)) +
  geom_histogram()
  
#summarize
newer_merged_data %>%
  filter(!is.na(SLPWEED)) %>%
  group_by(SLPWEED) %>%
    summarise(mean_waso = mean(waso),
              median_waso = median(waso))

```

```{r}
#WASO related to weed use before sleep

 waso_before_sleep <- glm(waso ~ SLPWEED + RACE + SEROSTAT + SMOKE + EMTWBR, data = newer_merged_data, family = "gaussian")

summary(waso_before_sleep)

waso_twopot <- glm(waso ~ twopotuse + RACE + SEROSTAT + SMOKE + EMTWBR, data = newer_merged_data, family = "gaussian")

summary(waso_twopot)
```
```{r}
# visits needed

newer_merged_data %>%
  group_by(VISIT) %>%
  count()


```
```{r}
newer_merged_data %>%
  filter(AGENOX < 40) %>%
  nrow()

```

### CESD
```{r}

#read in data

cesd <- read_csv("data_request_07192025.csv")

#merge in CESD
newest_merged_data <- newer_merged_data %>%
  left_join(cesd, by = c("MACSID", "VISIT"))

newest_merged_data %>%
  sum(is.na(CESD))

#264 missing is a fair amount

newer_merged_data %>%
  summarise(mean_twopot = mean(twopotuse.y))

newer_merged_data %>%
  filter(twopotuse.y > 0) %>%
  summarise(mean_twopot = mean(twopotuse.y))
```

### addressing missing data for AD use and CESD

for cesd, could impute with mean or use alternative visit when available. for antidepressant use, would have to either get MACS medication data or could use based on visit 65

## final analysis

need to take out smoking and add in age

```{r}

#redo categorical collapsing

almost_final_data <- newest_merged_data %>%
  mutate(RACE = as.factor(RACE),
         EDUCA2 = as.factor(EDUCA2)) %>%
  mutate(race_collapsed = fct_collapse(RACE,
                                       White = 1,
                                       Black = 3,
                                       Hispanic = c(2, 4, 8),
                                       Other = c(5, 6, 7)),
         schooling = fct_collapse(EDUCA2,
                                  'Less than HS' = c(1, 2),
                                  'HS or some college' = c(3, 4, 5),
                                  'College or more' = c(6, 7, 8)),
         sleep_aid_collapsed = fct_collapse(sleep_aid,
                                            'Used Sleep Aid' = c("OTC Sleep Aid", "Prescription sleep aid", "antihistamine", "benzo"),
                                            'Did not use' = "None"),
         cesd_dichot = ifelse(CESD < 16, 1, 2),
         cesd_factor = as.factor(cesd_dichot))

almost_final_data %>%
  count(race_collapsed)


#just 32 missing for slpweed in final data, should i just exclude those people?
```

```{r}
#create final dataset
nearly_final_data <- almost_final_data %>%
  select(MACSID, VISIT, AGENOX, SLPWEED, race_collapsed, SEROSTAT, schooling, sleep_aid_collapsed, CESD, cesd_factor, twopotuse.y, remlat, PSQIRATE, tstmin, seslponset, SMOKE, slplat) %>%
  rename(twopotuse = twopotuse.y) %>%
  #WASO
  mutate(spt = ((tstmin*100)/seslponset)+0.5,
         waso = spt-tstmin)
```

```{r}
#missing data in this dataset


str(nearly_final_data)


#clean variable types
nearly_final_data <- nearly_final_data %>%
  mutate(SEROSTAT = as.factor(SEROSTAT),
         SLPWEED = as.factor(SLPWEED),
         PSQIRATE = as.factor(PSQIRATE))

miss_var_summary(nearly_final_data)

```

```{r}
#take out people who don't have slpweed, don't want to impute that

nearly_final_data <- nearly_final_data %>%
  filter(!is.na(SLPWEED))
```

```{r}
#make sure things look good in this dataset
#remlat
summary(nearly_final_data$remlat)

nearly_final_data %>%
  ggplot(aes(x = remlat)) +
  geom_histogram()
```
Right skewed witha. few outliers

```{r}
#sleep latency
summary(nearly_final_data$slplat)

nearly_final_data %>%
  ggplot(aes(x = slplat)) +
  geom_histogram()
```
also right skewed with a few outliers

```{r}
#psqirate

summary(nearly_final_data$PSQIRATE)
```

### antidepressant data

```{r}
#read in new antidepressant data
file_path <- "/Users/danrzewnicki/Downloads/2026_02_24_X19024.xlsx"

med_data <- read_excel(file_path)

#inspect the data
str(med_data)

#narrow down to columns that I need
med_data <- med_data %>%
  select(MACSID, VISIT, NAME_ENTERED, UNK_DEPRESS)

#make visit values match
id_visit_lookup <- nearly_final_data %>%
  select(MACSID, VISIT) %>%
  mutate(VISIT = as.character(VISIT)) %>%
  mutate(VISIT = sub("0$", "", VISIT),
         VISIT = as.numeric(VISIT))

id_visit_lookup <- id_visit_lookup %>%
  mutate(MACSID = trimws(MACSID),
         VISIT = trimws(VISIT))

med_data <- med_data %>%
  mutate(MACSID = trimws(MACSID),
         VISIT = trimws(VISIT))

#right join to get the right ID-visit combos
antidepressants <- med_data %>%
  right_join(id_visit_lookup,  by = c("MACSID" = "MACSID", "VISIT" = "VISIT"))


```

```{r}
#identify antidepressants
#antidepressant use
ssri <- c("citalopram", "celexa", "escitalopram", "lexapro", "fluoxetine", "prozac", "paroxetine", "paxil", "sertraline", "zoloft")

snri <- c("desvenlafaxine", "pristiq", "duloxetine", "cymbalta", "venlafaxine", "effexor")

tca <- c("amitriptyline", "elavil", "clomipramine", "anafranil", "desipramine", "norpramine", "doxepin", "sinequan", "imipramine", "tofranil", "nortriptyline", "pamelor")

maoi <- c("isocarboxazid", "marplan", "phenelzine", "nardil", "tranylcypromine", "parnate", "selegiline", "emsam", "rasagiline", "azilect")

#make column uniform
antidepressants <- antidepressants %>%
  mutate(name_clean = str_to_lower(NAME_ENTERED),
         name_clean = str_trim(name_clean, side = "both"))

#all antidepressants
all_antidepressants <- c(ssri, snri, tca, maoi)


# Build a regex pattern from your antidepressant list
antidep_pattern <- paste(all_antidepressants, collapse = "|")

#identify if med is antidepressant

antidepressants <- antidepressants %>%
  mutate(
    # First handle NAs, then check for exact substring match
    antidep_flag = case_when(
      is.na(name_clean) ~ 0L,
      str_detect(name_clean, antidep_pattern) ~ 1L,
      # Fuzzy match as fallback for misspellings
      sapply(name_clean, function(x) {
        any(stringsim(x, all_antidepressants, method = "jw") >= 0.85)
      }) ~ 1L,
      TRUE ~ 0L
    )
  )



#summarise to one per ID
antidep_by_id <- antidepressants %>%
  group_by(MACSID) %>%
  summarise(on_antidepressant = as.integer(any(antidep_flag == 1L)))


sum(is.na(antidep_by_id$on_antidepressant))

antidep_by_id %>%
  group_by(on_antidepressant) %>%
  count()

```

```{r}
#create final data
final_data <- nearly_final_data %>%
  mutate(MACSID = as.character(MACSID)) %>%
  left_join(antidep_by_id)
```

###final data cleaning

questions for prior to analysis
- remove outliers from remlat and sleep latency
- impute and fit sleep latency and psqi rate
- remove those with missing rem latency and fit one more time

```{r}
#impute data for cesd

# Check which variables correlate with CESD
# Calculate correlations and sort by strength
numeric_data <- nearly_final_data[, sapply(nearly_final_data, is.numeric)]
correlations <- cor(numeric_data, use = "pairwise.complete.obs")[, "CESD"]
correlations <- sort(abs(correlations), decreasing = TRUE)
print(correlations)

#associations with categorical variables
#slpweed
ggplot(nearly_final_data, aes(x = SLPWEED, y = CESD)) +
  geom_boxplot()

#race
ggplot(nearly_final_data, aes(x = race_collapsed, y = CESD)) +
  geom_boxplot()

#serostat
ggplot(nearly_final_data, aes(x = SEROSTAT, y = CESD)) +
  geom_boxplot()



```


```{r}

#make new model function
final_model <- function(x, y) {
  model <- glm(x ~ SLPWEED + race_collapsed + SEROSTAT + AGENOX + schooling + on_antidepressant + cesd_factor, data = y, family = "gaussian")
}


```


```{r}
#final data demographics

final_data %>%
  summarise(mean_age = mean(AGENOX),
            proportion_white = (sum(race_collapsed == "White")/nrow(final_data)),
            mean_psqi = mean(PSQISCORE, na.rm = TRUE))

#proprotion of use by HIV status

chisq.test(final_data$SLPWEED, final_data$SEROSTAT)

table(final_data$SLPWEED, final_data$SEROSTAT)

sum(is.na(final_data$waso))

sum(is.na(final_data$remlat))

sum(is.na(final_data$PSQIRATE))

final_data

```

next things to do: 
- move model results function
- rename estimate, term, etc

```{r}
#remlatency
rem_data <- final_data %>%
  filter(!is.na(remlat))

rem_latency <- final_model(complete_rem$remlat, complete_rem)
rem_latency

export_model_results(rem_latency, "new_rem_latency_results")

export_model_results <- function(model_results, file_name, output_dir = getwd()) {

formatted_results <- function(outcome) { outcome %>%
  tidy_and_attach(conf.int = T) %>%
  tidy_add_reference_rows %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value)%>%
  filter(term != "(Intercept)")%>%
  mutate(term = case_when(term == "SLPWEED1" ~"Used Cannabis Prior to Sleep",
                          term == "SLPWEED2" ~ "   Yes",
                          term == "race_collapsedWhite" ~ "Race",
                          term == "race_collapsedHispanic"~ "   Hispanic",
                          term == "race_collapsedBlack" ~ "   Black", 
                          term == "race_collapsedOther" ~ "   Other",
                          term == "AGENOX" ~ "Age",
                          term == "SEROSTAT1" ~ "HIV Serostatus",
                          term == "SEROSTAT2" ~ "   HIV+",
                          term == "schoolingLess than HS" ~ "Education Level",
                          term == "schoolingHS or some college" ~ "   HS or some college",
                          term == "schoolingCollege or more" ~ "   College or more",
                          term == "on_antidepressant" ~ "Using antidepressants",
                          term == "cesd_factor2" ~ "   Yes",
                          term == "cesd_factor1" ~ "Depression",
                          T ~ term))%>%
  mutate(across(c(estimate:conf.high), ~round(., 2)))%>%
  mutate(p.value = round(p.value, 3),
         p.value = if_else(p.value <0.001, "<0.01",as.character(p.value)))%>%
  unite("CI",conf.low, conf.high, sep="-")%>%
  unite("estimate",estimate, CI, sep= " (")%>%
  mutate(estimate = str_c(estimate, ")")) %>%
   mutate(estimate = str_remove_all(estimate, fixed("NA (NA-NA)"))) %>%
  replace_na(list(p.value = "", estimate = "")) %>%
  rename(`expression(beta[1])` = estimate,
                  Coefficient = term)
 
}
formatted_results

# Build full file path
  file_path <- file.path(output_dir, paste0(file_name, ".csv"))
  output_dir <- getwd()
  
  # Write to CSV
  write.csv(formatted_results, file = file_path, row.names = FALSE)
}

  #mutate(term = factor(term, levels=c("Piperacillin-tazobactam",
                                      "Piperacillin-tazobactam:wbc","wbc",
                                      "charlson_score","sofa_max_24bcx","decade",
                                      "non_white","female","time_to_rx","First-stage residual"
```

```{r}
#sleep latency
sleep_latency <- final_model(complete_cases$slplat, complete_cases)

sleep_latency_results <- formatted_results(sleep_latency)

sleep_latency_results

```

```{r}
#waso
waso <- final_model(complete_cases$waso, complete_cases)

waso_results <- formatted_results(waso)

waso_results
```

```{r}
#psqi sleep quality

psqi <- final_model(final_data$PSQIRATE)
psqi
```


```{r}
#final data and twopot

#make twopot function
final_model_twopot <- function(x, y) {
  model <- glm(x ~ twopotuse + race_collapsed + factor(SEROSTAT) + AGENOX + schooling + on_antidepressant + cesd_factor, data = y, family = "gaussian")
}

rem_latency_twopot <- final_model_twopot(complete_rem$remlat, complete_rem)

twopot_remlat <- formatted_results(twopot_remlat)
twopot_remlat
```

```{r}
#twopot and waso

waso_twopot <- final_model_twopot(complete_cases$waso, complete_cases)
waso_twopot <- formatted_results(waso_twopot)
waso_twopot
```

```{r}
#sleep latency twopot
sleep_latency_twopot <- final_model_twopot(complete_cases$slplat, complete_cases)

sleep_latency_twopot <- formatted_results(sleep_latency_twopot)
sleep_latency_twopot
```

```{r}
psqi_twopot <- final_model_twopot(final_data$PSQIRATE)
psqi_twopot
```

```{r}
#export results
batch_export_model_results <- function(results_list, file_name, output_dir = getwd()) {
  
  # Combine all results with a model label column
  combined_results <- bind_rows(
    lapply(names(results_list), function(model_name) {
      results_list[[model_name]] %>%
        mutate(model = model_name) %>%
        select(model, everything())
    })
  ) %>%
    mutate(
      estimate     = round(estimate, 4),
      std.error    = round(std.error, 4),
      statistic    = round(statistic, 4),
      p.value      = round(p.value, 4),
      conf.low     = round(conf.low, 4),
      conf.high    = round(conf.high, 4),
      significance = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        p.value < 0.1   ~ ".",
        TRUE            ~ ""
      )
    )
  
  # Build full file path
  file_path <- file.path(output_dir, paste0(file_name, ".csv"))
  
  # Write to CSV
  write.csv(combined_results, file = file_path, row.names = FALSE)
  
  message("Batch results exported to: ", file_path)
  return(invisible(combined_results))
}

# Run all six models
results_list <- list(rem_latency, sleep_latency, waso, rem_latency_twopot, sleep_latency_twopot, waso_twopot)


# Batch export
batch_export_model_results(
  results_list = results_list,
  file_name    = "all_sleep_model_results"
)
```

```{r}
#data request

rzewnicki_request <- almost_final_data %>%
  select(MACSID, VISIT)

write.csv(rzewnicki_request, file = "rzewnicki_request.csv", row.names = FALSE)

rzewnicki_request %>%
  group_by(VISIT) %>%
  count()
```

```{r}
#inspect final data for missingness

miss_var_summary(final_data)

#only variable of interest with missing data is CES-D



#impute data in remlat dataset (don't want to impute outcomes)
install.packages("mice")
library(mice)

#remove cesd_factor
impute_data <- rem_data %>%
  select(-cesd_factor)

imp <- mice(impute_data, m = 5, seed = 123)

# Check imputation quality
plot(imp)  # Convergence
densityplot(imp)  # Distribution of imputations

# Create dichotomized version
imp_long <- complete(imp, "long", include = TRUE)
imp_long$CESD_high <- as.factor(ifelse(imp_long$CESD >= 16, 1, 0))  # Example cutoff
imp_final <- as.mids(imp_long)

# Run analysis
# Fit different models
fit1 <- with(imp_final, glm(remlat ~ SLPWEED + race_collapsed + SEROSTAT + AGENOX + schooling + on_antidepressant + CESD_high, family = "gaussian"))
fit2 <- with(imp_final, glm(slplat ~ SLPWEED + race_collapsed + SEROSTAT + AGENOX + schooling + on_antidepressant + CESD_high, family = "gaussian"))
fit3 <- with(imp_final, glm(waso ~ SLPWEED + race_collapsed + SEROSTAT + AGENOX + schooling + on_antidepressant + CESD_high, family = "gaussian"))

# Pool each
pooled1 <- pool(fit1)
pooled2 <- pool(fit2)
pooled3 <- pool(fit3)

# View results
summary(pooled1)
summary(pooled2)
summary(pooled3)

```
## reporting findings

Of note for the future: labelled package can be use to label variables with metadata that can be carried through to tables and figures with gtsummary 

```{r}
#label data

library(labelled)

final_data <- final_data %>%
  set_variable_labels(
    remlat = "REM Latency",
    slplat = "Sleep Latency",
    waso = "Wake After Sleep Onset",
    AGENOX = "Age",
    SLPWEED = "Cannabis Use Prior to Sleep",
    race_collapsed = "Race",
    SEROSTAT = "HIV Serostatus",
    schooling = "Highest Education Level",
    cesd_factor = "Depression",
    twopotuse = "Average Cannabis Use",
    on_antidepressant = "Using Antidepressants"
  ) %>%
  set_value_labels(
    SLPWEED = c(
      No = 1,
      Yes = 2
    ))
    race_collapsed = c(
      "White" = "White",
      "Hispanic" = "Hispanic",
      "Black" = "Black",
      "Other" = "Other"
    ),
    SEROSTAT = c(
      "HIV-" = "1",
      "HIV+" = "2"
    ),
    schooling = c(
      "Less than HS" = "Less than HS",
      "HS or some College" = "HS or some College",
      "College degree or more" = "College or more"
    ),
    cesd_factor = c(
      "Not Depressed" = "1",
      "Depressed" = "2"
    ))
    
   
```
```{r}

export_model_results <- function(model_results, output_dir = getwd()) {

formatted_results <- model_results %>%
  tidy_and_attach(conf.int = T) %>%
  tidy_add_reference_rows %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value)%>%
  filter(term != "(Intercept)")%>%
  mutate(term = case_when(term == "SLPWEED1" ~"Used Cannabis Prior to Sleep",
                          term == "SLPWEED2" ~ "   Yes",
                          term == "race_collapsedWhite" ~ "Race",
                          term == "race_collapsedHispanic"~ "   Hispanic",
                          term == "race_collapsedBlack" ~ "   Black", 
                          term == "race_collapsedOther" ~ "   Other",
                          term == "AGENOX" ~ "Age",
                          term == "SEROSTAT1" ~ "HIV Serostatus",
                          term == "SEROSTAT2" ~ "   HIV+",
                          term == "schoolingLess than HS" ~ "Education Level",
                          term == "schoolingHS or some college" ~ "   HS or some college",
                          term == "schoolingCollege or more" ~ "   College or more",
                          term == "on_antidepressant" ~ "Using antidepressants",
                          term == "cesd_factor2" ~ "   Yes",
                          term == "cesd_factor1" ~ "Depression",
                          T ~ term))%>%
  mutate(across(c(estimate:conf.high), ~round(., 2)))%>%
  mutate(p.value = round(p.value, 3),
         p.value = if_else(p.value <0.001, "<0.01",as.character(p.value)))%>%
  unite("CI",conf.low, conf.high, sep="-")%>%
  unite("estimate",estimate, CI, sep= " (")%>%
  mutate(estimate = str_c(estimate, ")")) %>%
   mutate(estimate = str_remove_all(estimate, fixed("NA (NA-NA)"))) %>%
  replace_na(list(p.value = "", estimate = "")) %>%
  rename(`expression(beta[1])` = estimate,
                  Coefficient = term)

#extract input name
input_name <- deparse(substitute(model_results))
 
# Build full file path
  file_path <- file.path(output_dir, paste0(input_name, ".csv"))

  
  # Write to CSV
  write.csv(formatted_results, file = file_path, row.names = FALSE)
  message("Results exported to: ", file_path)
  return(invisible(formatted_results))
}

results <- list(rem_latency, sleep_latency, waso, rem_latency_twopot, sleep_latency_twopot, waso_twopot)

lapply(results, export_model_results)

export_model_results(rem_latency)
export_model_results(sleep_latency)
export_model_results(waso)
export_model_results(rem_latency_twopot)
export_model_results(sleep_latency_twopot)
export_model_results(waso_twopot)
```

```{r}
#complete data

complete_cases <- final_data %>%
  filter(!is.na(cesd_factor))

miss_var_summary(complete_cases)

complete_rem <- complete_cases %>%
  filter(!is.na(remlat))
```

